---
title: "ggplot2_challenges_solutions"
author: "D-Lab and workshop participants"
date: "3/21/2021"
output: html_document
---

# Load the [heart dataset](http://search.r-project.org/library/kmed/html/heart.html)

Use the heart dataset for the following challenges. Import it and save it in a variable named `heart`. Make sure to review the data dictionary linked above!

```{r}
library(ggplot2)
heart <- read.csv("data/heart.csv")
names(heart)
```


# Challenge 1 

Create a gg histogram using the heart dataset. Save it in a variable named `A`

```{r}
A <- ggplot(heart, aes(x = chol)) +
  geom_histogram(binwidth = 6) +    
  labs(x = "Serum cholestorol in mg/dl",
       y = "Count",
       title = "Histogram of Serum Cholestorol (mg/dl)") +
  theme_minimal()

A
```

# Challenge 2 

Create gg boxplots using the heart dataset. Save this figure in a variable named `B`

```{r}
library(dplyr)
# treat sex as a factor instad of continuous variable for plotting
heart <- heart %>%
  mutate(sex  = as.factor(sex))

B <- ggplot(heart, aes(x = sex, y = chol, fill = sex)) + 
  geom_boxplot() +
  scale_x_discrete(labels = c("Female", "Male")) +      # change x labels
  scale_fill_discrete(name = "Biological Sex",          # change fill legend labels
                      labels = c("Female", "Male")) +
  labs(x = element_blank(),
       y = "Serum cholestoral in mg/dl",
       title = "Boxplot of Serum Cholesteral (mg/dl)",
       subtitle = "by Biological Sex") +
  theme_bw()

B
```

# Challenge 3

Create a gg scatterplot using the heart dataset and save it in a variable named `C`

```{r}
C <- ggplot(heart, aes(x = age, y = chol, 
                       color = as.factor(sex), 
                       shape = as.factor(sex))) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, lwd = 1) +
  labs(x = "Age",
       y = "Serum cholestoral in mg/dl",
       title = "Serum Cholesteral (mg/dl) by Biological Sex") +
  theme_bw() 

C
```

# Challenge 4

Create a gg lineplot using the heart dataset and save it in a variable named `D`

```{r}
# sometimes we will want to re-shape our data for plotting
heart_plot <- heart %>%
  mutate(sex = ifelse(sex == 0, "Female", "Male"),  # we can also change labels in the data instead of in ggplot
         slope = as.factor(slope)) %>%
  group_by(age, sex, slope) %>%
  summarize(chol = mean(chol))

D <- ggplot(heart_plot, aes(x = age, y = chol, color = sex)) +
  facet_wrap(~sex) +
  geom_line() +
  scale_color_manual(values=c("#CC6666", "#9999CC"), name = "Biological Sex") + 
  labs(x = "Age",
       y = "Serum cholestoral in mg/dl",
       title = "Serum Cholesteral (mg/dl) by Biological Sex") +
  theme_minimal()

D
```

# Challenge 5 

Use cowplot to create a compound figure named "compound_figure" that contains subplots A, B, C, and D that you created above. 

```{r}
library(cowplot)
compound = plot_grid(A, B, C, D, 
                     nrow = 2, ncol = 2,
                     scale = 0.85, 
                     labels = c("A)", "B)", "C)", "D)"))
compound
```
