---
title: "R Graphics with ggplot2"
subtitle: "Challenges + ggmap and the Google Maps Static API"
author: "Evan Muzzall"
date: "April 15, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load the heart dataset

Use the heart dataset for the following challenges. Import it and save it in a variable named `heart`

```{r}
## YOUR CODE HERE
```


# Challenge 1 

Create a gg histogram using the heart dataset. Save it in a variable named `A`

```{r}
## YOUR CODE HERE
```

# Challenge 2 

Create gg boxplots using the heart dataset. Save this figure in a variable named `B`

```{r}
## YOUR CODE HERE
```

# Challenge 3

Create a gg scatterplot using the heart dataset and save it in a varibale named `C`

```{r}
## YOUR CODE HERE
```

# Challenge 4

Create a lineplot using the heart dataset and save it in a variable named `D`

```{r}
## YOUR CODE HERE
```

# Challenge 5 

Use cowplot to create a compound figure named "compound_figure" that contains subplots A, B, C, and D that you created above. 

```{r}
## YOUR CODE HERE
```
