---
title: 
output:
  revealjs::revealjs_presentation:
    theme: serif
    transition: fade
    highlight: tango
    center: true
width: 1200

---

##R Graphics

UC Berkeley Library and D-Lab

Fall 2017

-------


###Plotting in R

There are two major* sets of tools for creating plots in R:

**Base** graphics, which come with all R installations as is 


**ggplot2**, a specialty package that we'll be using mostly





<small>*lattice is another but base and gpplot2 are by far most used</small>

-----



### Gapminder Dataset
Gapminder is a country-year dataset with information on life expectancy, population, and GDP per capita.

```{r }
dat <- read.csv("data/gapminder-FiveYearData.csv", stringsAsFactors = F)
head(dat)
```

------

### R base graphics

*Minimal* call takes the following form

plot(x=)

```{r fig.height = 4, fig.align="center"}

plot(x = dat$lifeExp) 
```


------

### R base graphics

*Basic* call takes the following form

plot(x=, y=)

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp)
```


----

### R base graphics

- The *type* argument tells R what shape to use in plot

- "p" = point/scatter plots (default plotting behavior)

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p")
```

----

### R base graphics

- "l" = line plots


```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="l") 

```


----

### R base graphics

- "b" -- both line and point plots

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="b") 
```

----

### R base graphics

* Certain plot types require different calls outside of the "type" argument

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
hist(x=dat$lifeExp)
```

----


### R base graphics

* *breaks* changes the start/end of bins

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
hist(x=dat$lifeExp, breaks=5)
```

----


### R base graphics

* *breaks* changes the start/end of bins

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
hist(x=dat$lifeExp, breaks=10)
```

----




### R base graphics

* density plots

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
age.density<-density(x=dat$lifeExp, na.rm=T)
plot(x=age.density)
```

----


### R base graphics

* density plots

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Plot the density object, bandwidth of 0.5
plot(x=density(x=dat$lifeExp, bw=.5, na.rm=T))
```

----


### R base graphics

* density plots

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Plot the density object, bandwidth of 0.5
plot(x=density(x=dat$lifeExp, bw=4, na.rm=T))
```

----




### R base graphics

**Your turn**

1. Make a histogram to examine the distribution of the population variable. 

```{r eval = F}

hist(x=, breaks=)

```



----


### R base graphics

* Labels

```{r, eval=F}
plot(x=, y=, type="", xlab="", ylab="", main="") 
```

----



### R base graphics

* Labels

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p", 
     xlab="GDP per cap", ylab="Life Expectancy", main="Life Expectancy ~ GDP") # Add labels for axes and overall plot
```

----




### R base graphics


* Axis and size scaling

```{r, eval=F}
plot(x=, y=, type="", xlim=, ylim=, cex=)
```

----

### R base graphics


* Axis and size scaling

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp)
```

----


### R base graphics

* Axis and size scaling

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, xlim = c(1000,20000)) 

```


----


### R base graphics

* Axis and size scaling

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, xlim = c(1000,20000), cex=2) 

```


----


### R base graphics

* Axis and size scaling

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, xlim = c(1000,20000), cex=0.5) 

```


----



### R base graphics

* Basic call with popular scaling arguments

```{r, eval=F}
plot(x=, y=, type="", col="", pch=, lty=, lwd=)
```


----



### R base graphics

* Colors
```{r}
colors() # View all elements of the color vector
colors()[179] # View specific element of the color vector
```

----


### R base graphics

* Colors

http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

http://research.stowers.org/mcm/efg/R/Color/Chart/

----




### R base graphics

* Colors

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p", col=colors()[145]) # or col="gold3"
```


----



### R base graphics

* Colors

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p", col="seagreen4") # or col=colors()[578]
```


----


### R base graphics

* Point Styles and Widths

[A Good Reference](http://www.endmemo.com/program/R/pic/pchsymbols.png)


```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change point style to crosses
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p", pch=3) 
```


----


### R base graphics

* Point Styles and Widths


```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change point style to filled squares
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p",pch=15) 
```


----



### R base graphics

* Point Styles and Widths


```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change point style to filled squares and increase point size to 3
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p",pch=15, cex=3) 
```

----



### R base graphics

* Point Styles and Widths

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change point style to "w"
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p", pch="w")
```

----




### R base graphics

* Line Styles and Widths

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Line plot with solid line
plot(x = dat$gdpPercap, y = dat$lifeExp, type="l", lty=1)
```

----



### R base graphics

* Line Styles and Widths

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Line plot with medium dashed line
plot(x = dat$gdpPercap, y = dat$lifeExp, type="l", lty=2)
```

----




### R base graphics

* Line Styles and Widths

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change line width to 2
plot(x = dat$gdpPercap, y = dat$lifeExp, type="l", lty=3, lwd=2)
```

----




### R base graphics

* Line Styles and Widths

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# Change line width to 10 and use dash-dot
plot(x = dat$gdpPercap, y = dat$lifeExp, type="l",  lty=4, lwd=10)
```

----



### R base graphics

* Annotations, reference lines, and legends

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# plot the line first
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p")
# now add the label
text(x=40000, y=50, labels="Evens Out", cex = .75)
```

----


### R base graphics

* Annotations, reference lines, and legends

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
# plot the line
plot(x = dat$gdpPercap, y = dat$lifeExp, type="p")
# now the guides
abline(v=40000, h=75, lty=2)
```

----


### R base graphics

**Your turn**

Make a scatterplot with population on the x axis and life expectancy on the y axis. Change the color to "peachpuff3" and the point symbol to "+"

```{r eval = F}

plot(x=, y=, type=, col=, pch=)

```

----

## ggplot2

- More elegant and compact code

- Aesthetically pleasing defaults

- Powerful for exploratory data analysis

- Follows a grammar like a language


----


## ggplot2

```{r, eval=FALSE}
ggplot(data=, aes(x=, y=), color=, size=,) + geom_xxxx()+geom_yyyy()
```

The *grammar* involves some basic components:

1. **Data**: a data.frame
2. **Aes**thetics: How your data are represented visually, aka "mapping". Which variables are shown on x, y axes, as well as color, size, shape, etc.
3. **Geom**etry: The geometric objects in a plot. points, lines, polygons, etc.

----

## ggplot2


```{r fig.height = 4, fig.width = 6, fig.align = "center"}
library(ggplot2)
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp)) + geom_point()
```

----


## ggplot2

By itself, the call to ggplot isn’t enough to draw a figure:

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
library(ggplot2)
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp)) 
```

----


## ggplot2


```{r eval= F}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp)) + geom_point()

```

equivalent to:

```{r eval = F}
my_plot <- ggplot(data = dat, aes(x = gdpPercap, y = lifeExp))
my_plot + geom_point()
```

----


## ggplot2


**Your Turn**

Modify the example so that the figure visualises how life expectancy has changed over time:

<small>Hint: the gapminder dataset has a column called "year", which should appear on the x-axis.</small>

```{r eval = F}

ggplot(data = , aes(x = , y = )) + geom_point()


```


## ggplot2


Anatomy of **aes**

<small>We’ve used the **aes** function to tell the scatterplot geom about the **x** and **y** locations of each point. Another **aes** property we can modify is the point **color** </small>

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + geom_point()
```


----

## ggplot2

Anatomy of **aes**

**Color** isn't the only aesthetic argument we can set to display variation in the data. We can also vary by shape, size, etc.

```{r, eval=FALSE}
ggplot(data=, aes(x=, y=, by =, color=, linetype=, shape=, size=))
```

----


## ggplot2

Anatomy of **aes**


```{r fig.height = 4, fig.width = 6, fig.align = "center" }
ggplot(data = dat, aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()

```

----



## ggplot2

Anatomy of **aes**


```{r fig.height = 4, fig.width = 6, fig.align = "center" }
ggplot(data = dat, aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line() + geom_point()

```

----


## ggplot2

Anatomy of **aes**


```{r fig.height = 4, fig.width = 6, fig.align = "center" }
ggplot(data = dat, aes(x=year, y=lifeExp, by=country)) + 
  geom_line(aes(color=continent)) + geom_point()
```


----



## ggplot2


**Your Turn**

Switch the order of the point and line layers from the previous example. What happened?


----


## ggplot2

Labels are their own layers in ggplot.

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
# add x and y axis labels
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point() + xlab("GDP per capita") + ylab("Life Expectancy") + ggtitle("My fancy graph")
```

----



## ggplot2

...so are scales

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
# limit x axis from 1,000 to 20,000
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
geom_point() + xlab("GDP per capita") + ylab("Life Expectancy") + ggtitle("My fancy graph") + xlim(1000, 20000)
```

----



## ggplot2

Transformations and Stats

<small>ggplot also makes it easy to overlay statistical models over the data. To demonstrate we’ll go back to an earlier example:</small>

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + geom_point()
```

----


## ggplot2

Transformations and Stats

<small>We can change the scale of units on the x axis using the **scale** functions. These control the mapping between the data values and visual values of an aesthetic.</small>

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + geom_point() + scale_x_log10()
```

----




## ggplot2

Transformations and Stats

<small>We can fit a simple relationship to the data by adding another layer, **geom_smooth**:
</small>

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp, color=continent)) + geom_point() + scale_x_log10() + geom_smooth(method="lm")
```

----






## ggplot2

Transformations and Stats

<small>Note that we 5 lines, one for each region, because the **color** option is the global **aes** function.. But if we move it, we get different restuls:
</small>

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp)) + geom_point(aes(color=continent)) + scale_x_log10() + geom_smooth(method="lm")
```

----



## ggplot2

Transformations and Stats


```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = gdpPercap, y = lifeExp)) + geom_point(aes(color=continent)) + scale_x_log10() + geom_smooth(method="lm", size = 1.5)
```

----



## ggplot2


**Your Turn**


Modify the color and size of the points on the point layer in the previous example so that they are fixed (i.e. not reflective of continent).

<small>Hint: do not use the **aes** function.</small>

-----


## ggplot2

Facets

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = year, y = lifeExp)) + geom_point() + facet_wrap( ~ continent)
```

----


## ggplot2

Putting it all together

<small>RStudio provides a really useful [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) of the different layers available, and more extensive documentation is available on the [ggplot2 website](http://docs.ggplot2.org/current/). Finally, if you have no idea how to change something, a quick google search will usually send you to a relevant question and answer on Stack Overflow with reusable code to modify!</small>


----



## ggplot2

Bar Plots

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = lifeExp)) + geom_bar(stat="bin")

```

----



## ggplot2

Bar Plots

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = lifeExp, fill = continent)) + geom_bar(stat="bin")

```

----



## ggplot2

Box Plots

```{r fig.height = 4, fig.width = 6, fig.align = "center"}
ggplot(data = dat, aes(x = continent, y = lifeExp)) + geom_boxplot()

```

----



## ggplot2


**Your Turn**


Create a density plot of GDP per capita, filled by continent.


<small>
Advanced: 
>Transform the x axis to better visualise the data spread. 
>Add a facet layer to panel the density plots by year.
</small>

-----



## ggplot2

Exporting- raster/bitmap

```{r eval=FALSE}
jpeg(filename="example.png", width=, height=)
plot(x,y)
dev.off()

```



Exporting- Vector
```{r eval=FALSE}
pdf(filename="example.pdf", width=, height=)
plot(x,y)
dev.off()

```

----

## ggplot2

Exporting with ggplot

```{r, eval=FALSE}
# Assume we saved our plot is an object called example.plot
ggsave(filename="example.pdf", plot=example.plot, scale=, width=, height=)
```


