---
title: "R Data Visualization Challenges"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load the [heart dataset](http://search.r-project.org/library/kmed/html/heart.html)

Use the heart dataset for the following challenges. Import it and save it in a variable named `heart`. Make sure to review the data dictionary linked above!

```{r import}
## YOUR CODE HERE
```

**For each challenge, use ggplot2**

# Challenge 1 

Using the heart dataset, create a histogram of the Serum cholesterol. Save it in a variable named `A`.

```{r challenge_1}
## YOUR CODE HERE
```

# Challenge 2 

Using the heart dataset, create a boxplot of Serum cholesterol based on patient self-described sex. Save this figure in a variable named `B`. 

Hint: The variable that identifies patient sex needs to be coerced into a different variable type in order to make this visualization.

```{r challenge_2}
## YOUR CODE HERE
```

# Challenge 3

Using the heart dataset, create a scatterplot that visualizes the relationship between patient age and Serum cholesterol. Save it in a variable named `C`

```{r challenge_3}
## YOUR CODE HERE
```

# Challenge 4

Create a line plot of the mean cholesterol for each age, separated by sex using the heart data. You will need to do some data manipulation with `dplyr` to make this plot. Save your plot to a variable named `D`.

```{r challenge_4}
## YOUR CODE HERE
```

# Challenge 5 

Use cowplot to create a compound figure named `compound_figure` that contains subplots `A`, `B`, `C`, and `D` that you created above. 

```{r challenge_5}
## YOUR CODE HERE
```
