---
title: "R Data Visualization Solutions"
output: html_document
---

# Load the [heart dataset](http://search.r-project.org/library/kmed/html/heart.html)

Use the heart dataset for the following challenges. Import it and save it in a variable named `heart`. Make sure to review the data dictionary linked above!

```{r import}
library(ggplot2)
heart <- read.csv("data/heart.csv")
names(heart)
```

**For each challenge, use** `ggplot2`

# Challenge 1 

Using the heart dataset, create a histogram of the Serum cholesterol. Save it in a variable named `A`.

```{r challenge_1}
A <- ggplot(heart, aes(x = chol)) +
  geom_histogram(binwidth = 6) +    
  labs(x = "Serum cholestorol in mg/dl",
       y = "Count",
       title = "Histogram of Serum Cholestorol (mg/dl)") +
  theme_minimal()

print(A)
```

# Challenge 2 

Using the heart dataset, create a boxplot of Serum cholesterol based on patient self-described sex. Save this figure in a variable named `B`. 

Hint: The variable that identifies patient sex needs to be coerced into a different variable type in order to make this visualization.

```{r challenge_2}
library(dplyr)
# treat sex as a factor instad of continuous variable for plotting
heart <- heart %>%
  mutate(sex  = as.factor(sex))

B <- ggplot(heart, aes(x = sex, y = chol, fill = sex)) + 
  geom_boxplot() +
  scale_x_discrete(labels = c("Female", "Male")) +      # change x labels
  scale_fill_discrete(name = "Biological Sex",          # change fill legend labels
                      labels = c("Female", "Male")) +
  labs(x = element_blank(),
       y = "Serum cholestoral in mg/dl",
       title = "Boxplot of Serum Cholesteral (mg/dl)",
       subtitle = "by Biological Sex") +
  theme_bw()

print(B)
```

# Challenge 3

Using the heart dataset, create a scatterplot that visualizes the relationship between patient age and Serum cholesterol. Save it in a variable named `C`

```{r challenge_3}
C <- ggplot(heart, aes(x = age, y = chol, 
                       color = as.factor(sex), 
                       shape = as.factor(sex))) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, lwd = 1) +
  labs(x = "Age",
       y = "Serum cholestoral in mg/dl",
       title = "Serum Cholesteral (mg/dl) by Biological Sex") +
  theme_bw() 

print(C)
```

# Challenge 4

Create a line plot of the mean cholesterol for each age, separated by sex using the heart data. You will need to do some data manipulation with `dplyr` to make this plot. Save your plot to a variable named `D`.

```{r challenge_4}
# Sometimes we will want to re-shape our data for plotting
heart_plot <- heart %>%
  mutate(sex = ifelse(sex == 0, "Female", "Male"),  # we can also change labels in the data instead of in ggplot
         slope = as.factor(slope)) %>%
  group_by(age, sex, slope) %>%
  summarize(chol = mean(chol))

D <- ggplot(heart_plot, aes(x = age, y = chol, color = sex)) +
  facet_wrap(~sex) +
  geom_line() +
  scale_color_manual(values=c("#CC6666", "#9999CC"), name = "Biological Sex") + 
  labs(x = "Age",
       y = "Serum cholestoral in mg/dl",
       title = "Serum Cholesteral (mg/dl) by Biological Sex") +
  theme_minimal()

print(D)
```

# Challenge 5 

Use cowplot to create a compound figure named `compound_figure` that contains subplots `A`, `B`, `C`, and `D` that you created above. 

```{r challenge_5}
library(cowplot)
compound = plot_grid(A, B, C, D, 
                     nrow = 2, ncol = 2,
                     scale = 0.85, 
                     labels = c("A)", "B)", "C)", "D)"))
print(compound)
```
